### 2013-12-06 paper.tex
First commit with sample input

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[runningheads,a4paper]{llncs}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}

\usepackage{url}
\urldef{\mailsa}\path|{alfred.hofmann, ursula.barth, ingrid.haas, frank.holzwarth,|
\urldef{\mailsb}\path|anna.kramer, leonie.kunz, christine.reiss, nicole.sator,|
\urldef{\mailsc}\path|erika.siebert-cole, peter.strasser, lncs}@springer.com|    
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\usepackage{epstopdf}

\begin{document}

\mainmatter  % start of an individual contribution

% first the title is needed
\title{Lecture Notes in Computer Science:\\Authors' Instructions
for the Preparation\\of Camera-Ready
Contributions\\to LNCS/LNAI/LNBI Proceedings}

% a short form should be given in case it is too long for the running head
\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Alfred Hofmann%
\thanks{Please note that the LNCS Editorial assumes that all authors have used
the western naming convention, with given names preceding surnames. This determines
the structure of the names in the running heads and the author index.}%
\and Ursula Barth\and Ingrid Haas\and Frank Holzwarth\and\\
Anna Kramer\and Leonie Kunz\and Christine Rei\ss\and\\
Nicole Sator\and Erika Siebert-Cole\and Peter Stra\ss er}
%
\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Springer-Verlag, Computer Science Editorial,\\
Tiergartenstr. 17, 69121 Heidelberg, Germany\\
\mailsa\\
\mailsb\\
\mailsc\\
\url{http://www.springer.com/lncs}}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

\toctitle{Lecture Notes in Computer Science}
\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}
The abstract should summarize the contents of the paper and should
contain at least 70 and at most 150 words. It should be written using the
\emph{abstract} environment.
\keywords{We would like to encourage you to list your keywords within
the abstract section}
\end{abstract}

\section{Introduction}

You are strongly encouraged to use \LaTeXe{} for the
preparation of your camera-ready manuscript together with the
corresponding Springer class file \verb+llncs.cls+. Only if you use
\LaTeXe{} can hyperlinks be generated in the online version
of your manuscript.

The \LaTeX{} source of this instruction file for \LaTeX{} users may be
used as a template. This is
located in the ``authors'' subdirectory in
\url{ftp://ftp.springer.de/pub/tex/latex/llncs/latex2e/instruct/} and
entitled \texttt{typeinst.tex}. There is a separate package for Word 
users. Kindly send the final and checked source
and PDF files of your paper to the Contact Volume Editor. This is
usually one of the organizers of the conference. You should make sure
that the \LaTeX{} and the PDF files are identical and correct and that
only one version of your paper is sent. It is not possible to update
files at a later stage. Please note that we do not need the printed
paper.

We would like to draw your attention to the fact that it is not possible
to modify a paper in any way, once it has been published. This applies
to both the printed book and the online version of the publication.
Every detail, including the order of the names of the authors, should
be checked before the paper is sent to the Volume Editors.

\subsection{Checking the PDF File}

Kindly assure that the Contact Volume Editor is given the name and email
address of the contact author for your paper. The Contact Volume Editor
uses these details to compile a list for our production department at
SPS in India. Once the files have been worked upon, SPS sends a copy of
the final pdf of each paper to its contact author. The contact author is
asked to check through the final pdf to make sure that no errors have
crept in during the transfer or preparation of the files. This should
not be seen as an opportunity to update or copyedit the papers, which is
not possible due to time constraints. Only errors introduced during the
preparation of the files will be corrected.

This round of checking takes place about two weeks after the files have
been sent to the Editorial by the Contact Volume Editor, i.e., roughly
seven weeks before the start of the conference for conference
proceedings, or seven weeks before the volume leaves the printer's, for
post-proceedings. If SPS does not receive a reply from a particular
contact author, within the timeframe given, then it is presumed that the
author has found no errors in the paper. The tight publication schedule
of LNCS does not allow SPS to send reminders or search for alternative
email addresses on the Internet.

In some cases, it is the Contact Volume Editor that checks all the final
pdfs. In such cases, the authors are not involved in the checking phase.

\subsection{Additional Information Required by the Volume Editor}

If you have more than one surname, please make sure that the Volume Editor
knows how you are to be listed in the author index.

\subsection{Copyright Forms}

The copyright form may be downloaded from the ``For Authors"
(Information for LNCS Authors) section of the LNCS Website:
\texttt{www.springer.com/lncs}. Please send your signed copyright form
to the Contact Volume Editor, either as a scanned pdf or by fax or by
courier. One author may sign on behalf of all of the other authors of a
particular paper. Digital signatures are acceptable.

\section{Paper Preparation}

Springer provides you with a complete integrated \LaTeX{} document class
(\texttt{llncs.cls}) for multi-author books such as those in the LNCS
series. Papers not complying with the LNCS style will be reformatted.
This can lead to an increase in the overall number of pages. We would
therefore urge you not to squash your paper.

Please always cancel any superfluous definitions that are
not actually used in your text. If you do not, these may conflict with
the definitions of the macro package, causing changes in the structure
of the text and leading to numerous mistakes in the proofs.

If you wonder what \LaTeX{} is and where it can be obtained, see the
``\textit{LaTeX project site}'' (\url{http://www.latex-project.org})
and especially the webpage ``\textit{How to get it}''
(\url{http://www.latex-project.org/ftp.html}) respectively.

When you use \LaTeX\ together with our document class file,
\texttt{llncs.cls},
your text is typeset automatically in Computer Modern Roman (CM) fonts.
Please do
\emph{not} change the preset fonts. If you have to use fonts other
than the preset fonts, kindly submit these with your files.

Please use the commands \verb+\label+ and \verb+\ref+ for
cross-references and the commands \verb+\bibitem+ and \verb+\cite+ for
references to the bibliography, to enable us to create hyperlinks at
these places.

For preparing your figures electronically and integrating them into
your source file we recommend using the standard \LaTeX{} \verb+graphics+ or
\verb+graphicx+ package. These provide the \verb+\includegraphics+ command.
In general, please refrain from using the \verb+\special+ command.

Remember to submit any further style files and
fonts you have used together with your source files.

\subsubsection{Headings.}

Headings should be capitalized
(i.e., nouns, verbs, and all other words
except articles, prepositions, and conjunctions should be set with an
initial capital) and should,
with the exception of the title, be aligned to the left.
Words joined by a hyphen are subject to a special rule. If the first
word can stand alone, the second word should be capitalized.

Here are some examples of headings: ``Criteria to Disprove
Context-Freeness of Collage Language", ``On Correcting the Intrusion of
Tracing Non-deterministic Programs by Software", ``A User-Friendly and
Extendable Data Distribution System", ``Multi-flip Networks:
Parallelizing GenSAT", ``Self-determinations of Man".

\subsubsection{Lemmas, Propositions, and Theorems.}

The numbers accorded to lemmas, propositions, and theorems, etc. should
appear in consecutive order, starting with Lemma 1, and not, for
example, with Lemma 11.

\subsection{Figures}

For \LaTeX\ users, we recommend using the \emph{graphics} or \emph{graphicx}
package and the \verb+\includegraphics+ command.

Please check that the lines in line drawings are not
interrupted and are of a constant width. Grids and details within the
figures must be clearly legible and may not be written one on top of
the other. Line drawings should have a resolution of at least 800 dpi
(preferably 1200 dpi). The lettering in figures should have a height of
2~mm (10-point type). Figures should be numbered and should have a
caption which should always be positioned \emph{under} the figures, in
contrast to the caption belonging to a table, which should always appear
\emph{above} the table; this is simply achieved as matter of sequence in
your source.

Please center the figures or your tabular material by using the \verb+\centering+
declaration. Short captions are centered by default between the margins
and typeset in 9-point type (Fig.~\ref{fig:example} shows an example).
The distance between text and figure is preset to be about 8~mm, the
distance between figure and caption about 6~mm.

To ensure that the reproduction of your illustrations is of a reasonable
quality, we advise against the use of shading. The contrast should be as
pronounced as possible.

If screenshots are necessary, please make sure that you are happy with
the print quality before you send the files.
\begin{figure}
\centering
\includegraphics[height=6.2cm]{eijkel2}
\caption{One kernel at $x_s$ (\emph{dotted kernel}) or two kernels at
$x_i$ and $x_j$ (\textit{left and right}) lead to the same summed estimate
at $x_s$. This shows a figure consisting of different types of
lines. Elements of the figure described in the caption should be set in
italics, in parentheses, as shown in this sample caption.}
\label{fig:example}
\end{figure}

Please define figures (and tables) as floating objects. Please avoid
using optional location parameters like ``\verb+[h]+" for ``here".

\paragraph{Remark 1.}

In the printed volumes, illustrations are generally black and white
(halftones), and only in exceptional cases, and if the author is
prepared to cover the extra cost for color reproduction, are colored
pictures accepted. Colored pictures are welcome in the electronic
version free of charge. If you send colored figures that are to be
printed in black and white, please make sure that they really are
legible in black and white. Some colors as well as the contrast of
converted colors show up very poorly when printed in black and white.

\subsection{Formulas}

Displayed equations or formulas are centered and set on a separate
line (with an extra line or halfline space above and below). Displayed
expressions should be numbered for reference. The numbers should be
consecutive within each section or within the contribution,
with numbers enclosed in parentheses and set on the right margin --
which is the default if you use the \emph{equation} environment, e.g.,
\begin{equation}
  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt \;  .
\end{equation}

Equations should be punctuated in the same way as ordinary
text but with a small space before the end punctuation mark.

\subsection{Footnotes}

The superscript numeral used to refer to a footnote appears in the text
either directly after the word to be discussed or -- in relation to a
phrase or a sentence -- following the punctuation sign (comma,
semicolon, or period). Footnotes should appear at the bottom of
the
normal text area, with a line of about 2~cm set
immediately above them.\footnote{The footnote numeral is set flush left
and the text follows with the usual word spacing.}

\subsection{Program Code}

Program listings or program commands in the text are normally set in
typewriter font, e.g., CMTT10 or Courier.

\medskip

\noindent
{\it Example of a Computer Program}
\begin{verbatim}
program Inflation (Output)
  {Assuming annual inflation rates of 7%, 8%, and 10%,...
   years};
   const
     MaxYears = 10;
   var
     Year: 0..MaxYears;
     Factor1, Factor2, Factor3: Real;
   begin
     Year := 0;
     Factor1 := 1.0; Factor2 := 1.0; Factor3 := 1.0;
     WriteLn('Year  7% 8% 10%'); WriteLn;
     repeat
       Year := Year + 1;
       Factor1 := Factor1 * 1.07;
       Factor2 := Factor2 * 1.08;
       Factor3 := Factor3 * 1.10;
       WriteLn(Year:5,Factor1:7:3,Factor2:7:3,Factor3:7:3)
     until Year = MaxYears
end.
\end{verbatim}
%
\noindent
{\small (Example from Jensen K., Wirth N. (1991) Pascal user manual and
report. Springer, New York)}

\subsection{Citations}

For citations in the text please use
square brackets and consecutive numbers: \cite{jour}, \cite{lncschap},
\cite{proceeding1} -- provided automatically
by \LaTeX 's \verb|\cite| \dots\verb|\bibitem| mechanism.

\subsection{Page Numbering and Running Heads}

There is no need to include page numbers. If your paper title is too
long to serve as a running head, it will be shortened. Your suggestion
as to how to shorten it would be most welcome.

\section{LNCS Online}

The online version of the volume will be available in LNCS Online.
Members of institutes subscribing to the Lecture Notes in Computer
Science series have access to all the pdfs of all the online
publications. Non-subscribers can only read as far as the abstracts. If
they try to go beyond this point, they are automatically asked, whether
they would like to order the pdf, and are given instructions as to how
to do so.

Please note that, if your email address is given in your paper,
it will also be included in the meta data of the online version.

\section{BibTeX Entries}

The correct BibTeX entries for the Lecture Notes in Computer Science
volumes can be found at the following Website shortly after the
publication of the book:
\url{http://www.informatik.uni-trier.de/~ley/db/journals/lncs.html}

\subsubsection*{Acknowledgments.} The heading should be treated as a
subsubsection heading and should not be assigned a number.

\section{The References Section}\label{references}

In order to permit cross referencing within LNCS-Online, and eventually
between different publishers and their online databases, LNCS will,
from now on, be standardizing the format of the references. This new
feature will increase the visibility of publications and facilitate
academic research considerably. Please base your references on the
examples below. References that don't adhere to this style will be
reformatted by Springer. You should therefore check your references
thoroughly when you receive the final pdf of your paper.
The reference section must be complete. You may not omit references.
Instructions as to where to find a fuller version of the references are
not permissible.

We only accept references written using the latin alphabet. If the title
of the book you are referring to is in Russian or Chinese, then please write
(in Russian) or (in Chinese) at the end of the transcript or translation
of the title.

The following section shows a sample reference list with entries for
journal articles \cite{jour}, an LNCS chapter \cite{lncschap}, a book
\cite{book}, proceedings without editors \cite{proceeding1} and
\cite{proceeding2}, as well as a URL \cite{url}.
Please note that proceedings published in LNCS are not cited with their
full titles, but with their acronyms!

\begin{thebibliography}{4}

\bibitem{jour} Smith, T.F., Waterman, M.S.: Identification of Common Molecular
Subsequences. J. Mol. Biol. 147, 195--197 (1981)

\bibitem{lncschap} May, P., Ehrlich, H.C., Steinke, T.: ZIB Structure Prediction Pipeline:
Composing a Complex Biological Workflow through Web Services. In: Nagel,
W.E., Walter, W.V., Lehner, W. (eds.) Euro-Par 2006. LNCS, vol. 4128,
pp. 1148--1158. Springer, Heidelberg (2006)

\bibitem{book} Foster, I., Kesselman, C.: The Grid: Blueprint for a New Computing
Infrastructure. Morgan Kaufmann, San Francisco (1999)

\bibitem{proceeding1} Czajkowski, K., Fitzgerald, S., Foster, I., Kesselman, C.: Grid
Information Services for Distributed Resource Sharing. In: 10th IEEE
International Symposium on High Performance Distributed Computing, pp.
181--184. IEEE Press, New York (2001)

\bibitem{proceeding2} Foster, I., Kesselman, C., Nick, J., Tuecke, S.: The Physiology of the
Grid: an Open Grid Services Architecture for Distributed Systems
Integration. Technical report, Global Grid Forum (2002)

\bibitem{url} National Center for Biotechnology Information, \url{http://www.ncbi.nlm.nih.gov}

\end{thebibliography}

\section*{Appendix: Springer-Author Discount}

LNCS authors are entitled to a 33.3\% discount off all Springer
publications. Before placing an order, the author should send an email, 
giving full details of his or her Springer publication,
to \url{orders-HD-individuals@springer.com} to obtain a so-called token. This token is a
number, which must be entered when placing an order via the Internet, in
order to obtain the discount.

\section{Checklist of Items to be Sent to Volume Editors}
Here is a checklist of everything the volume editor requires from you:

\begin{itemize}
\settowidth{\leftmargin}{{\Large$\square$}}\advance\leftmargin\labelsep
\itemsep8pt\relax
\renewcommand\labelitemi{{\lower1.5pt\hbox{\Large$\square$}}}

\item The final \LaTeX{} source files
\item A final PDF file
\item A copyright form, signed by one author on behalf of all of the
authors of the paper.
\item A readme giving the name and email address of the
corresponding author.
\end{itemize}
\end{document}

### 2013-12-06 paper.tex
Primeiras alterações feitas nos autores e título

@@
 \usepackage{graphicx}
+\usepackage[utf8]{inputenc}
 
 \usepackage{url}
-\urldef{\mailsa}\path|{alfred.hofmann, ursula.barth, ingrid.haas, frank.holzwarth,|
-\urldef{\mailsb}\path|anna.kramer, leonie.kunz, christine.reiss, nicole.sator,|
-\urldef{\mailsc}\path|erika.siebert-cole, peter.strasser, lncs}@springer.com|    
+\urldef{\mailsa}\path|{rodrigo.g.branco,dmbpaiva,istela}@gmail.com|    
 \newcommand{\keywords}[1]{\par\addvspace\baselineskip
@@
 % first the title is needed
-\title{Lecture Notes in Computer Science:\\Authors' Instructions
-for the Preparation\\of Camera-Ready
-Contributions\\to LNCS/LNAI/LNBI Proceedings}
+\title{Acessibilidade nas fases de Engenharia de Requisitos, Projeto e Codificação de Software: Uma ferramenta de apoio}
 
 % a short form should be given in case it is too long for the running head
-\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}
+
+%VERIFICAR
+\titlerunning{AccTrace}
 
@@
 %
-\author{Alfred Hofmann%
-\thanks{Please note that the LNCS Editorial assumes that all authors have used
-the western naming convention, with given names preceding surnames. This determines
-the structure of the names in the running heads and the author index.}%
-\and Ursula Barth\and Ingrid Haas\and Frank Holzwarth\and\\
-Anna Kramer\and Leonie Kunz\and Christine Rei\ss\and\\
-Nicole Sator\and Erika Siebert-Cole\and Peter Stra\ss er}
-%
-\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
+\author{Rodrigo Gonçalves de Branco%
+%\thanks{Please note that the LNCS Editorial assumes that all authors have used
+%the western naming convention, with given names preceding surnames. This determines
+%the structure of the names in the running heads and the author index.}%
+\and Débora Maria Barroso Paiva\and Maria Istela Cagnin}
+
+%VERIFICAR
+\authorrunning{AccTrace}
 % (feature abused for this document to repeat the title also on left hand pages)
@@
 % unless you accept that it will be published
-\institute{Springer-Verlag, Computer Science Editorial,\\
-Tiergartenstr. 17, 69121 Heidelberg, Germany\\
+\institute{Faculdade de Computação - FACOM, \\
+Universidade Federal de Mato Grosso do Sul - UFMS,\\
+Cidade Universitária, Campo Grande - MS, Brasil\\
 \mailsa\\
-\mailsb\\
-\mailsc\\
-\url{http://www.springer.com/lncs}}
+\url{http://facom.ufms.br/}}
 

### 2014-01-08 paper_backup.tex
.

--- 2013-12-06 paper.tex
@@
 %the structure of the names in the running heads and the author index.}%
-\and Débora Maria Barroso Paiva\and Maria Istela Cagnin}
+\and Débora Maria Barroso Paiva \and Maria Istela Cagnin}
 
@@
 % unless you accept that it will be published
-\institute{Faculdade de Computação - FACOM, \\
-Universidade Federal de Mato Grosso do Sul - UFMS,\\
+\institute{Faculdade de Computação - FACOM\\
+Universidade Federal de Mato Grosso do Sul - UFMS\\
 Cidade Universitária, Campo Grande - MS, Brasil\\
@@
 \emph{abstract} environment.
-\keywords{We would like to encourage you to list your keywords within
-the abstract section}
+\keywords{Accessibility, Requirement Traceability, Software Development
+Process, CASE Tool}
+%\keywords{We would like to encourage you to list your keywords within
+%the abstract section}
 \end{abstract}

### 2014-01-09 paper.tex
.

@@
 \emph{abstract} environment.
-\keywords{We would like to encourage you to list your keywords within
-the abstract section}
+\keywords{Accessibility, Requirement Traceability, Software Development
+Process, CASE Tool}
+%\keywords{We would like to encourage you to list your keywords within
+%the abstract section}
 \end{abstract}
 
-\section{Introduction}
+\section{Introdução}
 
-You are strongly encouraged to use \LaTeXe{} for the
-preparation of your camera-ready manuscript together with the
-corresponding Springer class file \verb+llncs.cls+. Only if you use
-\LaTeXe{} can hyperlinks be generated in the online version
-of your manuscript.
+A Internet se consolidou como um dos principais meios de comunicação, disseminação da informação e fornecimento de serviços nos dias atuais, sendo que determinados tipos de informações são disponibilizados apenas através deste meio.
+Fornecer o acesso a estas informações para uma imensa variedade de dispositivos, com tecnologias e especificações diferentes, se tornou um desafio.
+E, principalmente, pessoas com necessidades especiais, devido a suas características inerentes, necessitam de produtos especialmente desenvolvidos para este fim.
 
-The \LaTeX{} source of this instruction file for \LaTeX{} users may be
-used as a template. This is
-located in the ``authors'' subdirectory in
-\url{ftp://ftp.springer.de/pub/tex/latex/llncs/latex2e/instruct/} and
-entitled \texttt{typeinst.tex}. There is a separate package for Word 
-users. Kindly send the final and checked source
-and PDF files of your paper to the Contact Volume Editor. This is
-usually one of the organizers of the conference. You should make sure
-that the \LaTeX{} and the PDF files are identical and correct and that
-only one version of your paper is sent. It is not possible to update
-files at a later stage. Please note that we do not need the printed
-paper.
+Construir produtos acessíveis não é uma tarefa fácil e este assunto é alvo de muitas pesquisas \cite{lazar:04,brajnik:06,zeng:05}, existindo propostas específicas para integrar usabilidade e acessibilidade aos processos de Engenharia de Software \cite{springerlink:10.1007/978-3-642-02713-0,maia:10}.
+Contudo, devido a fatores como a recente exposição do tema nos meios de comunicação e a deficiência no treinamento e formação  dos desenvolvedores, muitos deles sequer sabem como codificar para tornar seus produtos acessíveis \cite{kavcic:05,alves:11}.
 
-We would like to draw your attention to the fact that it is not possible
-to modify a paper in any way, once it has been published. This applies
-to both the printed book and the online version of the publication.
-Every detail, including the order of the names of the authors, should
-be checked before the paper is sent to the Volume Editors.
+A utilização de ferramentas que apóiam os profissionais na construção de boas soluções acessíveis é muito comum, pois estas aumentam a produtividade e diminuem o esforço necessário.
+Estas ferramentas podem variar entre ferramentas CASE, IDEs, frameworks, simuladores, validadores, avaliadores e até mesmo uma mistura destas citadas.
+Contudo, pesquisas mostram que as ferramentas disponíveis deixam a desejar no quesito de apoio à acessibilidade \cite{trewin:10}.
 
-\subsection{Checking the PDF File}
+A situação é agravada pela constatação de que, apesar de existirem vários estudos sobre rastreabilidade de requisitos genéricos \cite{ali:11,gotel:94,soonsongtanee:2010,mader:12}, poucos estudos têm indicado como ocorre a evolução dos requisitos de acessibilidade durante o processo de desenvolvimento das aplicações web \cite{dias:10}, e como os desenvolvedores utilizarão essa informação para construir o produto.
 
-Kindly assure that the Contact Volume Editor is given the name and email
-address of the contact author for your paper. The Contact Volume Editor
-uses these details to compile a list for our production department at
-SPS in India. Once the files have been worked upon, SPS sends a copy of
-the final pdf of each paper to its contact author. The contact author is
-asked to check through the final pdf to make sure that no errors have
-crept in during the transfer or preparation of the files. This should
-not be seen as an opportunity to update or copyedit the papers, which is
-not possible due to time constraints. Only errors introduced during the
-preparation of the files will be corrected.
-
-This round of checking takes place about two weeks after the files have
-been sent to the Editorial by the Contact Volume Editor, i.e., roughly
-seven weeks before the start of the conference for conference
-proceedings, or seven weeks before the volume leaves the printer's, for
-post-proceedings. If SPS does not receive a reply from a particular
-contact author, within the timeframe given, then it is presumed that the
-author has found no errors in the paper. The tight publication schedule
-of LNCS does not allow SPS to send reminders or search for alternative
-email addresses on the Internet.
-
-In some cases, it is the Contact Volume Editor that checks all the final
-pdfs. In such cases, the authors are not involved in the checking phase.
-
-\subsection{Additional Information Required by the Volume Editor}
-
-If you have more than one surname, please make sure that the Volume Editor
-knows how you are to be listed in the author index.
-
-\subsection{Copyright Forms}
-
-The copyright form may be downloaded from the ``For Authors"
-(Information for LNCS Authors) section of the LNCS Website:
-\texttt{www.springer.com/lncs}. Please send your signed copyright form
-to the Contact Volume Editor, either as a scanned pdf or by fax or by
-courier. One author may sign on behalf of all of the other authors of a
-particular paper. Digital signatures are acceptable.
+Este artigo apresenta o AccTrace, uma ferramenta CASE desenvolvida como um plugin do Eclipse para, através da rastreabilidade dos requisitos de acessibilidade, entregar ao desenvolvedor informações relevantes para a construção do produto.
 
@@
 
+\bibitem{lazar:04} Lazar, J., Dudley-Sponaugle, A.,  Greenidge, K.-D.: Improving web accessibility:
+a study of webmaster perceptions. Computers in Human Behavior, 20(2), 269--288. (2004)
+
+\bibitem{brajnik:06} Brajnik, G.: Web Accessibility Testing: When the Method Is the Culprit.
+In: Miesenberger, K., Klaus, J., Zagler, W.
+L., e Karshmer, A.
+I.
+(eds.) ICCHP 2006, vol.
+4061, pp.
+156--163.
+Springer (2006)
+
+\bibitem{zeng:05} Parmanto, B. e Zeng, X.: Metric for Web accessibility evaluation. JASIST, 56(13), 1394--1404. (2005)
+
+\bibitem{springerlink:10.1007/978-3-642-02713-0} Moreno, L., Martínez, P., Ruiz-Mezcua, B.: Integrating HCI in a Web Accessibility Engineering Approach.
+In: Stephanidis, C., (ed.) Universal Access in Human-Computer Interaction.
+Applications and Services, vol.
+5616, pp.
+745--754.
+Springer (2009)
+
+\bibitem{maia:10} Maia, L. S.: Um processo para o desenvolvimento de aplicações Web Acessíveis.
+Master Thesis, UFMS (2010)
+
+\bibitem{kavcic:05} Kavcic, A.: Software Accessibility: Recommendations and Guidelines. In: Computer as a Tool. EUROCON, The International Conference, vol. 2, pp 1024--1027. (2005)
+
+\bibitem{alves:11} Alves, D. D.: Acessibilidade no Desenvolvimento de Software Livre. Master Thesis, UFMS (2011)
+
+\bibitem{trewin:10} Trewin, S., Cragun, B., Swart, C., Brezin, J., Richards, J.: Accessibility challenges
+and tool features: an IBM Web developer perspective. In: Proceedings of the
+2010 International Cross Disciplinary Conference on Web Accessibility. W4A 2010, pp 32:1--32:10. ACM. (2010)
+
+\bibitem{dias:10} Dias, A. L., de Mattos Fortes, R. P., Masiero, P. C., Goularte, R.: Uma Revisão
+Sistemática sobre a inserção de Acessibilidade nas fases de desenvolvimento da Engenharia
+de Software em sistemas Web. In: Proceedings of the IX Symposium on Human
+Factors in Computing Systems. IHC 2010, pp. 39--48. SBC (2010)
+
+\bibitem{ali:11} Ali, N., Gueheneuc, Y., Antoniol, G.: Trust-Based Requirements Traceability.
+In Program Comprehension. ICPC 2011, pp. 111--120. IEEE (2011)
+
+\bibitem{gotel:94} Gotel, O. C. Z., Finkelstein, A. C. W.: An analysis of the requirements traceability
+problem. In Requirements Engineering, Proceedings of the First International Conference on, pp. 94--101 (1994)
+
+\bibitem{soonsongtanee:2010} Soonsongtanee, S., Limpiyakorn, Y.: Enhancement of requirements traceability
+with state diagrams. In: Computer Engineering and Technology (ICCET), 2nd International Conference on, vol. 2, pp. 248--252 (2010)
+
+\bibitem{mader:12} Mader, P., Egyed, A.: Assessing the effect of requirements traceability for software
+maintenance. In: Software Maintenance (ICSM), 28th IEEE International Conference on, pp. 171--180 (2012)
+
 \bibitem{jour} Smith, T.F., Waterman, M.S.: Identification of Common Molecular

### 2014-01-09 paper.tex
.

@@
 % first the title is needed
-\title{Acessibilidade nas fases de Engenharia de Requisitos, Projeto e Codificação de Software: Uma ferramenta de apoio}
+\title{AccTrace: Acessibilidade nas fases de Engenharia de Requisitos, Projeto e Codificação de Software}
 
@@
 
-Este artigo apresenta o AccTrace, uma ferramenta CASE desenvolvida como um plugin do Eclipse para, através da rastreabilidade dos requisitos de acessibilidade, entregar ao desenvolvedor informações relevantes para a construção do produto.
+Este artigo apresenta o AccTrace, uma ferramenta CASE desenvolvida como um plugin do Eclipse para, através da rastreabilidade dos requisitos de acessibilidade, entregar ao desenvolvedor informações relevantes para a construção de um produto acessível.
+O AccTrace foi baseado em um processo de desenvolvimento de softwares que incluía tarefas de acessibilidades baseado na ISO 12207, chamado MTA \cite{maia:10}.
 
@@
 
+\bibitem{maia:10} Maia, L. S.: Um processo para o desenvolvimento de aplicações Web Acessíveis.
+Master Thesis, UFMS (2010)
+
 \bibitem{jour} Smith, T.F., Waterman, M.S.: Identification of Common Molecular

### 2014-01-12 paper.tex
.

@@
 Construir produtos acessíveis não é uma tarefa fácil e este assunto é alvo de muitas pesquisas \cite{lazar:04,brajnik:06,zeng:05}, existindo propostas específicas para integrar usabilidade e acessibilidade aos processos de Engenharia de Software \cite{springerlink:10.1007/978-3-642-02713-0,maia:10}.
-Contudo, devido a fatores como a recente exposição do tema nos meios de comunicação e a deficiência no treinamento e formação  dos desenvolvedores, muitos deles sequer sabem como codificar para tornar seus produtos acessíveis \cite{kavcic:05,alves:11}.
+Contudo, devido a fatores como a recente exposição do tema nos meios de comunicação e a deficiência no treinamento e formação  dos desenvolvedores, muitos deles sequer sabem como codificar para tornar seus produtos acessíveis \cite{1630123,alves:11}.
 
@@
 Estas ferramentas podem variar entre ferramentas CASE, IDEs, frameworks, simuladores, validadores, avaliadores e até mesmo uma mistura destas citadas.
-Contudo, pesquisas mostram que as ferramentas disponíveis deixam a desejar no quesito de apoio à acessibilidade \cite{trewin:10}.
+Contudo, pesquisas mostram que as ferramentas disponíveis deixam a desejar no quesito de apoio à acessibilidade \cite{Trewin:2010:ACT:1805986.1806029}.
 
-A situação é agravada pela constatação de que, apesar de existirem vários estudos sobre rastreabilidade de requisitos genéricos \cite{ali:11,gotel:94,soonsongtanee:2010,mader:12}, poucos estudos têm indicado como ocorre a evolução dos requisitos de acessibilidade durante o processo de desenvolvimento das aplicações web \cite{dias:10}, e como os desenvolvedores utilizarão essa informação para construir o produto.
+A situação é agravada pela constatação de que, apesar de existirem vários estudos sobre rastreabilidade de requisitos genéricos \cite{5970169,292398,5485417,6405269}, poucos estudos têm indicado como ocorre a evolução dos requisitos de acessibilidade durante o processo de desenvolvimento das aplicações web \cite{analuizadias:2010}, e como os desenvolvedores utilizarão essa informação para construir o produto.
 
-Este artigo apresenta o AccTrace, uma ferramenta CASE desenvolvida como um plugin do Eclipse para, através da rastreabilidade dos requisitos de acessibilidade, entregar ao desenvolvedor informações relevantes para a construção de um produto acessível.
-O AccTrace foi baseado em um processo de desenvolvimento de softwares que incluía tarefas de acessibilidades baseado na ISO 12207, chamado MTA \cite{maia:10}.
+Este artigo apresenta o AccTrace, uma ferramenta CASE desenvolvida como um
+plugin do Eclipse para, através da rastreabilidade dos requisitos de
+acessibilidade, entregar ao desenvolvedor informações relevantes para a
+construção de um produto acessível. O AccTrace foi baseado em um processo de
+desenvolvimento de softwares que incluía tarefas de acessibilidades baseado na
+ISO 12207, chamado MTA \cite{maia:10}. Durante o processo de desenvolvimento do
+software, as referências entre requisitos e modelos UML, juntamente com a nova
+abordagem de especificar quais serão as técnicas de implementação de
+acessibilidade para estes relacionamentos, resultam em um comentário
+personalizado no código fonte, que é recuperado em tempo real detalhando as
+informações descritas anteriormente.
 
-\section{Paper Preparation}
+\subsection{Exemplo Motivacional}
 
-Springer provides you with a complete integrated \LaTeX{} document class
-(\texttt{llncs.cls}) for multi-author books such as those in the LNCS
-series. Papers not complying with the LNCS style will be reformatted.
-This can lead to an increase in the overall number of pages. We would
-therefore urge you not to squash your paper.
+A rastreabilidade dos requisitos geralmente é feita utilizando matrizes de
+rastreabilidade \cite{guo:2009:OBI:1681515.1682933}. Contudo, nas abordagens tradicionais, os
+requisitos de acessibilidade não são diferenciados dos outros requisitos. Desta
+forma, a implementação de tais requisitos não se torna explícita.
+Desenvolvedores experientes podem não encontrar problemas na implementação, mas
+desenvolvedores iniciantes podem não ter o mesmo sucesso.
 
-Please always cancel any superfluous definitions that are
-not actually used in your text. If you do not, these may conflict with
-the definitions of the macro package, causing changes in the structure
-of the text and leading to numerous mistakes in the proofs.
+É certo que o documento WCAG 2.0 pode e deve ser usado como documento de
+referência, pois ele contém critérios de sucesso, técnicas para implementação,
+etc. Contudo, este documento não é de fácil leitura para iniciantes,
+principalmente no que tange encontrar a informação necessária. Por este motivo,
+inclusive, que normalmente o documento é usado para avaliar e validar produtos
+prontos.
 
-If you wonder what \LaTeX{} is and where it can be obtained, see the
-``\textit{LaTeX project site}'' (\url{http://www.latex-project.org})
-and especially the webpage ``\textit{How to get it}''
-(\url{http://www.latex-project.org/ftp.html}) respectively.
+O desenvolvedor seria beneficiado caso a informação de o que deve ser feito, no
+que tange acessibilidade, estivesse disponível. Isto é exatamente o que nossa
+abordagem tenta resolver.
 
-When you use \LaTeX\ together with our document class file,
-\texttt{llncs.cls},
-your text is typeset automatically in Computer Modern Roman (CM) fonts.
-Please do
-\emph{not} change the preset fonts. If you have to use fonts other
-than the preset fonts, kindly submit these with your files.
+\section{Trabalhos Relacionados}
 
-Please use the commands \verb+\label+ and \verb+\ref+ for
-cross-references and the commands \verb+\bibitem+ and \verb+\cite+ for
-references to the bibliography, to enable us to create hyperlinks at
-these places.
-
-For preparing your figures electronically and integrating them into
-your source file we recommend using the standard \LaTeX{} \verb+graphics+ or
-\verb+graphicx+ package. These provide the \verb+\includegraphics+ command.
-In general, please refrain from using the \verb+\special+ command.
-
-Remember to submit any further style files and
-fonts you have used together with your source files.
-
-\subsubsection{Headings.}
-
-Headings should be capitalized
-(i.e., nouns, verbs, and all other words
-except articles, prepositions, and conjunctions should be set with an
-initial capital) and should,
-with the exception of the title, be aligned to the left.
-Words joined by a hyphen are subject to a special rule. If the first
-word can stand alone, the second word should be capitalized.
-
-Here are some examples of headings: ``Criteria to Disprove
-Context-Freeness of Collage Language", ``On Correcting the Intrusion of
-Tracing Non-deterministic Programs by Software", ``A User-Friendly and
-Extendable Data Distribution System", ``Multi-flip Networks:
-Parallelizing GenSAT", ``Self-determinations of Man".
-
-\subsubsection{Lemmas, Propositions, and Theorems.}
-
-The numbers accorded to lemmas, propositions, and theorems, etc. should
-appear in consecutive order, starting with Lemma 1, and not, for
-example, with Lemma 11.
-
-\subsection{Figures}
-
-For \LaTeX\ users, we recommend using the \emph{graphics} or \emph{graphicx}
-package and the \verb+\includegraphics+ command.
-
-Please check that the lines in line drawings are not
-interrupted and are of a constant width. Grids and details within the
-figures must be clearly legible and may not be written one on top of
-the other. Line drawings should have a resolution of at least 800 dpi
-(preferably 1200 dpi). The lettering in figures should have a height of
-2~mm (10-point type). Figures should be numbered and should have a
-caption which should always be positioned \emph{under} the figures, in
-contrast to the caption belonging to a table, which should always appear
-\emph{above} the table; this is simply achieved as matter of sequence in
-your source.
-
-Please center the figures or your tabular material by using the \verb+\centering+
-declaration. Short captions are centered by default between the margins
-and typeset in 9-point type (Fig.~\ref{fig:example} shows an example).
-The distance between text and figure is preset to be about 8~mm, the
-distance between figure and caption about 6~mm.
-
-To ensure that the reproduction of your illustrations is of a reasonable
-quality, we advise against the use of shading. The contrast should be as
-pronounced as possible.
-
-If screenshots are necessary, please make sure that you are happy with
-the print quality before you send the files.
-\begin{figure}
-\centering
-\includegraphics[height=6.2cm]{eijkel2}
-\caption{One kernel at $x_s$ (\emph{dotted kernel}) or two kernels at
-$x_i$ and $x_j$ (\textit{left and right}) lead to the same summed estimate
-at $x_s$. This shows a figure consisting of different types of
-lines. Elements of the figure described in the caption should be set in
-italics, in parentheses, as shown in this sample caption.}
-\label{fig:example}
-\end{figure}
-
-Please define figures (and tables) as floating objects. Please avoid
-using optional location parameters like ``\verb+[h]+" for ``here".
-
-\paragraph{Remark 1.}
-
-In the printed volumes, illustrations are generally black and white
-(halftones), and only in exceptional cases, and if the author is
-prepared to cover the extra cost for color reproduction, are colored
-pictures accepted. Colored pictures are welcome in the electronic
-version free of charge. If you send colored figures that are to be
-printed in black and white, please make sure that they really are
-legible in black and white. Some colors as well as the contrast of
-converted colors show up very poorly when printed in black and white.
-
-\subsection{Formulas}
-
-Displayed equations or formulas are centered and set on a separate
-line (with an extra line or halfline space above and below). Displayed
-expressions should be numbered for reference. The numbers should be
-consecutive within each section or within the contribution,
-with numbers enclosed in parentheses and set on the right margin --
-which is the default if you use the \emph{equation} environment, e.g.,
-\begin{equation}
-  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
-  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt \;  .
-\end{equation}
-
-Equations should be punctuated in the same way as ordinary
-text but with a small space before the end punctuation mark.
-
-\subsection{Footnotes}
-
-The superscript numeral used to refer to a footnote appears in the text
-either directly after the word to be discussed or -- in relation to a
-phrase or a sentence -- following the punctuation sign (comma,
-semicolon, or period). Footnotes should appear at the bottom of
-the
-normal text area, with a line of about 2~cm set
-immediately above them.\footnote{The footnote numeral is set flush left
-and the text follows with the usual word spacing.}
-
-\subsection{Program Code}
-
-Program listings or program commands in the text are normally set in
-typewriter font, e.g., CMTT10 or Courier.
-
-\medskip
-
-\noindent
-{\it Example of a Computer Program}
-\begin{verbatim}
-program Inflation (Output)
-  {Assuming annual inflation rates of 7%, 8%, and 10%,...
-   years};
-   const
-     MaxYears = 10;
-   var
-     Year: 0..MaxYears;
-     Factor1, Factor2, Factor3: Real;
-   begin
-     Year := 0;
-     Factor1 := 1.0; Factor2 := 1.0; Factor3 := 1.0;
-     WriteLn('Year  7% 8% 10%'); WriteLn;
-     repeat
-       Year := Year + 1;
-       Factor1 := Factor1 * 1.07;
-       Factor2 := Factor2 * 1.08;
-       Factor3 := Factor3 * 1.10;
-       WriteLn(Year:5,Factor1:7:3,Factor2:7:3,Factor3:7:3)
-     until Year = MaxYears
-end.
-\end{verbatim}
-%
-\noindent
-{\small (Example from Jensen K., Wirth N. (1991) Pascal user manual and
-report. Springer, New York)}
-
-\subsection{Citations}
-
-For citations in the text please use
-square brackets and consecutive numbers: \cite{jour}, \cite{lncschap},
-\cite{proceeding1} -- provided automatically
-by \LaTeX 's \verb|\cite| \dots\verb|\bibitem| mechanism.
-
-\subsection{Page Numbering and Running Heads}
-
-There is no need to include page numbers. If your paper title is too
-long to serve as a running head, it will be shortened. Your suggestion
-as to how to shorten it would be most welcome.
-
-\section{LNCS Online}
-
-The online version of the volume will be available in LNCS Online.
-Members of institutes subscribing to the Lecture Notes in Computer
-Science series have access to all the pdfs of all the online
-publications. Non-subscribers can only read as far as the abstracts. If
-they try to go beyond this point, they are automatically asked, whether
-they would like to order the pdf, and are given instructions as to how
-to do so.
-
-Please note that, if your email address is given in your paper,
-it will also be included in the meta data of the online version.
-
-\section{BibTeX Entries}
-
-The correct BibTeX entries for the Lecture Notes in Computer Science
-volumes can be found at the following Website shortly after the
-publication of the book:
-\url{http://www.informatik.uni-trier.de/~ley/db/journals/lncs.html}
-
-\subsubsection*{Acknowledgments.} The heading should be treated as a
-subsubsection heading and should not be assigned a number.
-
-\section{The References Section}\label{references}
-
-In order to permit cross referencing within LNCS-Online, and eventually
-between different publishers and their online databases, LNCS will,
-from now on, be standardizing the format of the references. This new
-feature will increase the visibility of publications and facilitate
-academic research considerably. Please base your references on the
-examples below. References that don't adhere to this style will be
-reformatted by Springer. You should therefore check your references
-thoroughly when you receive the final pdf of your paper.
-The reference section must be complete. You may not omit references.
-Instructions as to where to find a fuller version of the references are
-not permissible.
-
-We only accept references written using the latin alphabet. If the title
-of the book you are referring to is in Russian or Chinese, then please write
-(in Russian) or (in Chinese) at the end of the transcript or translation
-of the title.
-
-The following section shows a sample reference list with entries for
-journal articles \cite{jour}, an LNCS chapter \cite{lncschap}, a book
-\cite{book}, proceedings without editors \cite{proceeding1} and
-\cite{proceeding2}, as well as a URL \cite{url}.
-Please note that proceedings published in LNCS are not cited with their
-full titles, but with their acronyms!
+O tema Acessibilidade no Processo de desenvolvimento de software ainda
+é pouco explorado, com poucos trabalhos publicados a respeito
+\cite{maia:10,5599835}. Em contrapartida, podem ser encontrados diversos
+estudos que dizem respeito ao rastreamento de requisitos
+\cite{5970169,292398,5485417,6405269}, bem como encontrar estudos que utilizam
+ontologias para o mapeamento do domínio no processo de desenvolvimento ou na
+rastreabilidade dos requisitos \cite{5223183,6511842,4148940,5362244}. Contudo,
+não foi encontrado na literatura estudos específicos sobre a rastreabilidade de
+requisitos de acessibilidade durante o processo de desenvolvimento de software.
 
@@
 
-\bibitem{kavcic:05} Kavcic, A.: Software Accessibility: Recommendations and Guidelines. In: Computer as a Tool. EUROCON, The International Conference, vol. 2, pp 1024--1027. (2005)
+\bibitem{1630123} Kavcic, A.: Software Accessibility: Recommendations and Guidelines. In: Computer as a Tool. EUROCON, The International Conference, vol. 2, pp 1024--1027. (2005)
 
@@
 
-\bibitem{trewin:10} Trewin, S., Cragun, B., Swart, C., Brezin, J., Richards, J.: Accessibility challenges
+\bibitem{Trewin:2010:ACT:1805986.1806029} Trewin, S., Cragun, B., Swart, C., Brezin, J., Richards, J.: Accessibility challenges
 and tool features: an IBM Web developer perspective. In: Proceedings of the
@@
 
-\bibitem{dias:10} Dias, A. L., de Mattos Fortes, R. P., Masiero, P. C., Goularte, R.: Uma Revisão
+\bibitem{analuizadias:2010} Dias, A. L., de Mattos Fortes, R. P., Masiero, P. C., Goularte, R.: Uma Revisão
 Sistemática sobre a inserção de Acessibilidade nas fases de desenvolvimento da Engenharia
@@
 
-\bibitem{ali:11} Ali, N., Gueheneuc, Y., Antoniol, G.: Trust-Based Requirements Traceability.
+\bibitem{5970169} Ali, N., Gueheneuc, Y., Antoniol, G.: Trust-Based Requirements Traceability.
 In Program Comprehension. ICPC 2011, pp. 111--120. IEEE (2011)
 
-\bibitem{gotel:94} Gotel, O. C. Z., Finkelstein, A. C. W.: An analysis of the requirements traceability
-problem. In Requirements Engineering, Proceedings of the First International Conference on, pp. 94--101 (1994)
+\bibitem{292398} Gotel, O. C. Z., Finkelstein, A. C. W.: An analysis of the requirements traceability
+problem. In: Requirements Engineering, Proceedings of the First International
+Conference on, pp. 94--101 (1994)
 
-\bibitem{soonsongtanee:2010} Soonsongtanee, S., Limpiyakorn, Y.: Enhancement of requirements traceability
+\bibitem{5485417} Soonsongtanee, S., Limpiyakorn, Y.: Enhancement of requirements traceability
 with state diagrams. In: Computer Engineering and Technology (ICCET), 2nd International Conference on, vol. 2, pp. 248--252 (2010)
 
-\bibitem{mader:12} Mader, P., Egyed, A.: Assessing the effect of requirements traceability for software
+\bibitem{6405269} Mader, P., Egyed, A.: Assessing the effect of requirements traceability for software
 maintenance. In: Software Maintenance (ICSM), 28th IEEE International Conference on, pp. 171--180 (2012)
 
-\bibitem{maia:10} Maia, L. S.: Um processo para o desenvolvimento de aplicações Web Acessíveis.
-Master Thesis, UFMS (2010)
+\bibitem{guo:2009:OBI:1681515.1682933} guo, Y., Yang, M., Wang, J., Yang, P., Li, F.: An Ontology
+Based Improved Software Requirement Traceability Matrix. In: Proceedings of the
+2009 Second International Symposium on Knowledge Acquisition and Modeling,
+Vol. 01, pp. 160--163. IEEE (2009)
 
-\bibitem{jour} Smith, T.F., Waterman, M.S.: Identification of Common Molecular
-Subsequences. J. Mol. Biol. 147, 195--197 (1981)
+\bibitem{5599835} Moulin, C., Sbodio, M.L.: Improving the accessibility and
+efficiency of e-Government processes. In: Cognitive Informatics (ICCI), 2010 9th
+IEEE International Conference on, pp. 603--610 (2010)
 
-\bibitem{lncschap} May, P., Ehrlich, H.C., Steinke, T.: ZIB Structure Prediction Pipeline:
-Composing a Complex Biological Workflow through Web Services. In: Nagel,
-W.E., Walter, W.V., Lehner, W. (eds.) Euro-Par 2006. LNCS, vol. 4128,
-pp. 1148--1158. Springer, Heidelberg (2006)
+\bibitem{5223183} Assawamekin, N., Sunetnanta, T., Pluempitiwiriyawej, C.: MUPRET: An
+Ontology-Driven Traceability Tool for Multiperspective Requirements Artifacts.
+In: Computer and Information Science, 2009. ICIS 2009. Eighth IEEE/ACIS
+International Conference on, pp. 943--948 (2009)
 
-\bibitem{book} Foster, I., Kesselman, C.: The Grid: Blueprint for a New Computing
-Infrastructure. Morgan Kaufmann, San Francisco (1999)
+\bibitem{6511842} Martins, J., Machado, R.: Ontologies for Product and Process
+Traceability at Manufacturing Organizations: A Software Requirements Approach.
+In: Quality of Information and Communications Technology (QUATIC), 2012 Eighth
+International Conference on the, pp. 353--358 (2012)
 
-\bibitem{proceeding1} Czajkowski, K., Fitzgerald, S., Foster, I., Kesselman, C.: Grid
-Information Services for Distributed Resource Sharing. In: 10th IEEE
-International Symposium on High Performance Distributed Computing, pp.
-181--184. IEEE Press, New York (2001)
+\bibitem{4148940} Noll, R., Ribeiro, M.: Ontological Traceability over
+the Unified Process. In: Engineering of Computer-Based Systems, 2007. ECBS '07.
+14th Annual IEEE International Conference and Workshops on the, pp.
+249--255 (2007)
 
-\bibitem{proceeding2} Foster, I., Kesselman, C., Nick, J., Tuecke, S.: The Physiology of the
-Grid: an Open Grid Services Architecture for Distributed Systems
-Integration. Technical report, Global Grid Forum (2002)
-
-\bibitem{url} National Center for Biotechnology Information, \url{http://www.ncbi.nlm.nih.gov}
+\bibitem{5362244} guo, Y., Yang, M., Wang, J., Yang, P., Li, F.: An Ontology Based Improved
+Software Requirement Traceability Matrix. In: Knowledge Acquisition and
+Modeling, 2009. KAM '09. Second International Symposium on, vol. 1, pp.
+160--163 (2009)
 
@@
 
-\section*{Appendix: Springer-Author Discount}
-
-LNCS authors are entitled to a 33.3\% discount off all Springer
-publications. Before placing an order, the author should send an email, 
-giving full details of his or her Springer publication,
-to \url{orders-HD-individuals@springer.com} to obtain a so-called token. This token is a
-number, which must be entered when placing an order via the Internet, in
-order to obtain the discount.
-
-\section{Checklist of Items to be Sent to Volume Editors}
-Here is a checklist of everything the volume editor requires from you:
-
-\begin{itemize}
-\settowidth{\leftmargin}{{\Large$\square$}}\advance\leftmargin\labelsep
-\itemsep8pt\relax
-\renewcommand\labelitemi{{\lower1.5pt\hbox{\Large$\square$}}}
-
-\item The final \LaTeX{} source files
-\item A final PDF file
-\item A copyright form, signed by one author on behalf of all of the
-authors of the paper.
-\item A readme giving the name and email address of the
-corresponding author.
-\end{itemize}
 \end{document}

### 2014-01-12 paper.tex
.

@@
 
+Os trabalhos anteriores relacionados a este podem ser divididos em três grupos:
+Acessibilidade no Processo de Desenvolvimento de Software, Rastreabilidade de
+Requisitos e Ontologias para o Mapeamento do Domínio.
+
+\subsection{Acessibilidade no Processo de Desenvolvimento de Software}
+
 O tema Acessibilidade no Processo de desenvolvimento de software ainda
-é pouco explorado, com poucos trabalhos publicados a respeito
-\cite{maia:10,5599835}. Em contrapartida, podem ser encontrados diversos
+é pouco explorado. Maia \cite{maia:10} propôs  integrar usabilidade e
+acessibilidade aos processos de Engenharia de Software criando o MTA (Modelo de
+Tarefas de Acessibilidade), um modelo de desenvolvimento, baseado na norma ISO/IEC 12207, aplicável
+ao WCAG, que permite o apoio ferramental ao modelo.
+
+\subsection{Rastreabilidade de Requisitos}
+
+Diversos estudos que dizem respeito ao rastreamento de requisitos podem ser
+encontrados na literatura.
+
+ Em contrapartida, podem ser encontrados diversos
 estudos que dizem respeito ao rastreamento de requisitos
-\cite{5970169,292398,5485417,6405269}, bem como encontrar estudos que utilizam
+\cite{5970169,292398,5485417,6405269},
+
+\subsection{Ontologias para o Mapeamento do Domínio}
+
+bem como encontrar estudos que utilizam
 ontologias para o mapeamento do domínio no processo de desenvolvimento ou na

### 2014-01-14 paper.tex
.

@@
 
+\section{Abordagem}
+
+O AccTrace utiliza como escopo o Processo de Desenvolvimento de Software do MTA
+\cite{maia:10}, e os subprocessos 4 (Análise de Requisitos do Software), 5
+(Projeto de Software) e 6 (Construção do software). Como o escopo deste trabalho
+diz respeito a rastreabilidade dos requisitos de acessibilidade e suas técnicas
+de implementaçÕ, as tarefas de testes, presentes no Subprocesso 6, não serão
+levadas em consideração.
+
+A figura \ref{fig:figmagica} apresenta a abordagem escolhida e mostra como,
+partindo da Engenharia de Requisitos e chegando na construção do software,
+a rastreabilidade dos requisitos e das técnicas de implementação de
+acessibilidade é alcançada.
+
+\begin{figure}
+\centering
+\includegraphics[height=6.2cm]{img/figuramagica.png}
+\caption{Detalhamento dos Subprocessos do MTA para prover a rastreabilidade dos
+requisitos de acessibilidade de acordo com a abordagem adotada neste trabalho}
+\label{fig:figmagica}
+\end{figure}
+
+\subsection{Especialista em Acessibilidade}
+
+O MTA apresenta o papel de Especialista em Acessibilidade, fundamental em todas
+as etapas das tarefas de acessibilidade. Assim como o MTA, o AccTrace supõe que
+a equipe de desenvolvimento possui tal persona, cujo papel é identificar os
+requisitos de acessibilidade, bem como apresentar a associação dos requisitos,
+modelos UML e técnicas de implementação de acessibilidade, que serão explicadas
+adiante.
+
+\subsection{Subprocesso 4 - Análise de Requisitos de Software}
+
+O Subprocesso 4 traz duas tarefas de acessibilidade (\textbf{Estabelecer e
+Documentar os Requisitos de Acessibilidade do Software} e \textbf{Avaliar os
+Requisitos de Acessibilidade do Software}). A saída da tarefa {Avaliar os
+Requisitos de Acessibilidade do Software} gera o artefato \textbf{Documento de
+requisitos de acessibilidade do software avaliados e aprovados}. Neste trabalho,
+propõe-se que este artefato seja gerado em um formato legível por máquina. O
+formato escolhido neste trabalho foi o XMI/XML.
+
+Matrizes de rastreabilidade são usadas para promover a rastreabilidade dos
+requisitos de acessibilidade
+
 \begin{thebibliography}{4}

### 2014-01-14 paper.tex
.

@@
 
-A figura \ref{fig:figmagica} apresenta a abordagem escolhida e mostra como,
+Neste trabalho, propõe-se que os artefatos gerados pelo MTA aqui utilizados e os
+artefatos novos introduzidos estejam em um formato legível por máquina, para
+persistência de tais informações no modelo de dados da ferramenta.
+
+A Figura \ref{fig:figmagica} apresenta a abordagem escolhida e mostra como,
 partindo da Engenharia de Requisitos e chegando na construção do software,
@@
 Requisitos de Acessibilidade do Software} gera o artefato \textbf{Documento de
-requisitos de acessibilidade do software avaliados e aprovados}. Neste trabalho,
-propõe-se que este artefato seja gerado em um formato legível por máquina. O
-formato escolhido neste trabalho foi o XMI/XML.
+requisitos de acessibilidade do software avaliados e aprovados}.
 
-Matrizes de rastreabilidade são usadas para promover a rastreabilidade dos
-requisitos de acessibilidade
+A construção dos modelos UML na etapa Análise e Modelagem, conforme apresentado
+na Figura \ref{fig:figmagica}, recebe como entrada o artefato \textbf{Documento
+de Requisitos do software avaliado e aprovado}, para que a associação dos
+requisitos e modelos seja efetuada.
+
+\subsection{Subprocesso 5 - Projeto de Software}
+
+O Subprocesso 5 introduz três tarefas de acessibilidade (\textbf{Projetar
+Interfaces Externas Acessíveis}, \textbf{Realizar Projeto Navigacional
+Acessível} e \textbf{Avaliar a Acessibilidade do Projeto de Software}). O
+Documento de Requisitos e os Modelos UML são utilizados para realizar a
+especificação de Interfaces e do Projeto Navegacional, gerando o artefato
+\textbf{Projeto de Software Acessível}. Neste momento, é possível especificar
+quais serão as técnicas de implementação de acessibilidade e associá-las aos
+requisitos e modelos UML.
+
+\subsection{Subprocesso 6 - Construção do Software}
+
+O Subprocesso 6 possui duas tarefas importantes para o escopo deste trabalho
+(\textbf{Especificar Técnicas para Implementação da Acessibilidade da
+Interface e do Código} e \textbf{Codificar e documentar cada unidade de software
+de acordo com as técnicas de acessibilidade}). Para a rastreabilidade, a
+especificação das técnicas de implementação de acessibilidade já foram feitas no
+Subprocesso 5, faltando aqui efetuar a documentação para as mesmas. A etapa
+\textbf{Construção do Software} (Figura \ref{fig:figmagica}) utilizará os
+artefatos \textbf{Projeto de Software Acessível} e as especificações de técnicas
+de implementação de acessibilidade para finalmente permitir a construção do
+Software acessível.
+
+\section{Detalhes de Implementação}
 

### 2014-01-14 paper.tex
.

@@
 
-\begin{figure}
+\begin{figure}[h]
 \centering
@@
 \section{Detalhes de Implementação}
+
+O AccTrace foi desenvolvido utilizando as seguintes tecnologias:
+
+\begin{itemize}
+  \item MTA;
+  \item Eclipse Juno - IDE;
+  \item Requirement Designer v0.8.0 - Gerenciamento de Requisitos;
+  \item UML Designer v2.1.0 - Modelagem UML;
+  \item UML to Java Generator v1.0.2 - Geração de código;
+  \item Java JRE e JDK 1.7 - Desenvolvimento da ferramenta;
+  \item Projeto Aegis - ontologia de mapeamento do domínio de acessibilidade (usando o WCAG 2.0).
+\end{itemize}
+
+O Eclipse é uma plataforma madura e foi escolhido como IDE por vários motivos. Ele
+serve como base para diversos produtos e tecnologias baseadas em uma IDE, provendo
+uma API para facilitar a integração. O desenvolvimento de plugins para o Eclipse é feito diretamente na IDE,
+de forma prática e transparente, tendo ampla documentação a respeito. A linguagem
+utilizada para o desenvolvimento do plugin é a linguagem Java, assim como o código
+gerado pelo plugin de exportação.
+
+Os plugins de gerenciamento de requisitos (Requirement Designer) e modelagem (UML Designer) são desenvolvidos pela Obeo e são interoperáveis, portanto, para permitir uma integração suave entre as ferramentas, o plugin escolhido para geração de código foi o UML to Java Generator, desenvolvido pela
+mesma empresa.
+Todos estes plugins, assim como o AccTrace, persistem seus dados como arquivos RDF modelados com o formato EMF.
+Os plugins são disponibilizados sob a licença Eclipse Public License v1.0, assim como a própria IDE.
+Dessa forma, é possível estudar, alterar e customizar seus componentes para atingir os objetivos do trabalho.
+
+O Projeto Aegis \cite{aegis:13} define uma ontologia de acessibilidade, descrita no padrão OWL 1.0, e vai além de realizar o mapeamento dos conceitos de acessibilidade e cenários.
+Os arquivos disponibilizados mapeiam o modelo WCAG 2.0 (incluindo técnicas de implementação, critérios de sucesso e falha, etc), leitores de tela, navegadores textuais, lentes de aumento, WAI/ARIA, entre outros.
+
+A Figura \ref{fig:association} mostra o comportamento e relacionamento das ferramentas, tecnologias
+e atores envolvidos no desenvolvimento do produto.
+
+\begin{figure}[h]
+\centering
+\includegraphics[height=6.2cm]{./img/developmentNew2.png}
+\caption{Associação das ferramentas e atores no contexto do trabalho}
+\label{fig:association}
+\end{figure}
+
+Podemos interpretar a Figura \ref{fig:association} da seguinte forma:
+
+\begin{itemize}
+  \item O MTA é o processo de desenvolvimento e deve permear todas as fases do processo;
+  \item O especialista em acessibilidade (Accessibility Expert) deve participar das fases de
+desenvolvimento especificadas no MTA;
+  \item Os requisitos devem ser coletados e informados na fase de engenharia de requisitos
+(utilizando a ferramenta Requirement Designer) e os modelos e artefatos UML
+devem ser gerados (utilizando a ferramenta UML Designer). O especialista em
+acessibilidade deve auxiliar a alimentação destes dados, filltrando os requisitos de
+acessibilidade para que eles sejam associados aos modelos UML (utilizando a ferramenta
+Requirement Designer);
+  \item O AccTrace recupera as associações dos requisitos e modelos UML, permitindo que o especialista especifique as técnicas de implementaçãoo de acessibilidade para cada uma das associações.
+As técnicas, diretrizes, abordagens, dentre outros estão armazenadas em arquivos OWL embutidos na ferramenta;
+  \item A ferramenta de geração de código produzirá o código stub a partir dos modelos
+UML, adicionando as referências às associações de acessibilidade quando necessário;
+  \item Os desenvolvedores refinam o cóodigo até que o mesmo esteja apto a se tornar o
+produto acessível que será entregue.
+\end{itemize}
+
+A ferramenta possui três visões principais, de acordo com a Figura \ref{fig:acctrace}.
+No editor (AccTrace Editor) é possível alterar os repositórios dos requisitos e gerar as associações entre os modelos UML, requisitos e técnicas de implementação.
+Na visão
+dos requisitos (Requirement Associations) é possível visualizar quais requisitos associados
+ao modelo UML foram selecionados no editor. Na visão das técnicas já vinculadas
+(Accessibility Specifications View) é possível visualizar as técnicas de implementação já
+associadas, de acordo com o modelo UML selecionado no editor e o requisito de acessibilidade
+selecionado na visão dos requisitos. Além disso, é possível remover as técnicas de
+implementação associadas. As três visões são importantes para o correto funcionamento
+da ferramenta.
+
+\begin{figure}[h]
+\centering
+\includegraphics[height=6.2cm]{./img/acctrace.png}
+\caption{Visualização da ferramenta AccTrace na tela principal do
+Eclipse}
+\label{fig:acctrace}
+\end{figure}
+
+Uma vez selecionado o modelo UML e o requisito, é possível efetuar a associação da
+técnica de implementação de acessibilidade, clicando com o botãoo direito do mouse em
+cima do modelo UML, conforme demonstrado na Figura \ref{fig:rightclick}.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.45]{./img/rightclick.png}
+\caption{Procedimento para efetuar a associação da técnica de implementação de
+acessibilidade}
+\label{fig:rightclick}
+\end{figure}
+
+A hieraquia dos elementos da ontologia, presentes nos arquivos OWL, são apresentados em alto nível na Figura \ref{fig:ontologyrelationship}.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.3]{./img/ontologyrelationship.png}
+\caption{Relacionamento entre os elementos da ontologia (Elementos retirados
+da ontologia)}
+\label{fig:ontologyrelationship}
+\end{figure}
+
+Após os requisitos, modelos UML e as técnicas de implementação serem associadas, é possível gerar a matriz de rastreabilidade.
+Neste trabalho foi utilizado o Apache ODF Toolkit para geração do documento de rastreamento com formato Open-Document Spreadsheet file format (ODS), através da API de alto nível Simple.
+No documento, são geradas três planilhas: Requisitos \textit{versus} Modelos, Requisitos \textit{versus} Técnicas e Modelos \textit{versus} Técnicas.
+A planilha Requisitos \textit{versus} Modelos lista uma matriz de todos os Requisitos associados aos respectivos modelos.
+Requisitos que não estão associados a um modelo também são listados.
+Essa situação está prevista
+justamente para identificar problemas na associação que devam ser corrigidos. Para as
+planilhas Requisitos \textit{versus} Técnicas e Modelos \textit{versus} Técnicas, apenas os objetos que
+estão efetivamente referenciados na ferramenta são listados.
+
+Por fim, é possível efetuar a geração de código utilizando o plugin UML to Java Generator, que por sua vez utiliza o ponto de extensão do plugin Acceleo.
+O plugin UML to Java Generator foi especialmente modificado para que reconhecesse as associações entre Requisitos, Modelos UML e técnicas de implementação de acessibilidade, gerando comentários personalizados nos escopos corretos das classes de acordo com a expressão regular Java demonstrada a seguir:
+
+\medskip
+
+\noindent
+\begin{verbatim}
+String regex = "//!ACCTRACE!(/)?([^/\\\\0#]+(/)?)+#([^\\*\\*/])+";
+\end{verbatim}
+%
+\noindent
 
@@
 
+\bibitem{aegis:13} AEGIS Ontology, \url{http://www.aegis-project.eu/index.php?option=com_content&view=article&id=107&Itemid=65}
+
 \end{thebibliography}

### 2014-01-14 paper.tex
.

@@
 
+A expressão regular foi construída de modo a permitir, através de uma string pré-definida, a recuperação do arquivo Acctrace,
+bem como do elemento de associação referenciado. O comentário personalizado foi projetado de forma a não interfirir em outros
+comentários da linguagem. Iniciando pelas barras duplas (//), foi definido um
+identificador que tem a intenção de ser único (!ACCTRACE!), e o próprio ID do recurso RDF possui o caracter ``\#'', que é usado para separador entre
+o recurso (arquivo a ser aberto) e o elemento a ser resgatado (no caso, a
+associação criada entre o requisito e o modelo UML). A opção de se
+utilizar um comentário de linha (ao invés de comentários de bloco, usados por exemplo pela documentação JavaDoc) decorreu do fato
+do comentário descrito ser incluído dentro do corpo da classe e métodos, e caso
+o programador decida comentar o trecho de código inteiro, esse comentário não
+atrapalharia essa operação.
+
+A Figura \ref{fig:commentrecovery} mostra os passos necessários para recuração das informações de um comentário AccTrace, e a Figura \ref{fig:commentview}
+mostra a visão das informações recuperadas na visão específica para este fim no Eclipse.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=0.3]{./img/commentrecovery.png}
+\caption{Passos para recuperação das informações relevantes através de um
+comentário padrão AccTrace}
+\label{fig:commentrecovery}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.5]{./img/commentview.png}
+\caption{Explicitação do comentário selecionado}
+\label{fig:commentview}
+\end{figure}
+
 \begin{thebibliography}{4}

### 2014-01-14 paper.tex
.

@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \centering
-\includegraphics[height=6.2cm]{img/figuramagica.png}
+\includegraphics[scale=0.25]{img/figuramagica.png}
 \caption{Detalhamento dos Subprocessos do MTA para prover a rastreabilidade dos
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \centering
-\includegraphics[height=6.2cm]{./img/developmentNew2.png}
+\includegraphics[scale=0.25]{./img/developmentNew2.png}
 \caption{Associação das ferramentas e atores no contexto do trabalho}
@@
 A ferramenta possui três visões principais, de acordo com a Figura \ref{fig:acctrace}.
-No editor (AccTrace Editor) é possível alterar os repositórios dos requisitos e gerar as associações entre os modelos UML, requisitos e técnicas de implementação.
-Na visão
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=0.26]{./img/acctrace.png}
+\caption{Visualização da ferramenta AccTrace na tela principal do
+Eclipse}
+\label{fig:acctrace}
+\end{figure}
+
+No editor (AccTrace Editor) é possível alterar os repositórios dos requisitos e gerar as associações entre os modelos UML, requisitos e técnicas de implementação. Na visão
 dos requisitos (Requirement Associations) é possível visualizar quais requisitos associados
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[height=6.2cm]{./img/acctrace.png}
-\caption{Visualização da ferramenta AccTrace na tela principal do
-Eclipse}
-\label{fig:acctrace}
-\end{figure}
-
 Uma vez selecionado o modelo UML e o requisito, é possível efetuar a associação da
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \centering
-\includegraphics[scale=0.45]{./img/rightclick.png}
+\includegraphics[scale=0.25]{./img/rightclick.png}
 \caption{Procedimento para efetuar a associação da técnica de implementação de
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \centering
-\includegraphics[scale=0.3]{./img/ontologyrelationship.png}
+\includegraphics[scale=0.18]{./img/ontologyrelationship.png}
 \caption{Relacionamento entre os elementos da ontologia (Elementos retirados
@@
 \centering
-\includegraphics[scale=0.3]{./img/commentrecovery.png}
+\includegraphics[scale=0.17]{./img/commentrecovery.png}
 \caption{Passos para recuperação das informações relevantes através de um
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \centering
-\includegraphics[scale=0.5]{./img/commentview.png}
+\includegraphics[scale=0.45]{./img/commentview.png}
 \caption{Explicitação do comentário selecionado}
@@
 \end{figure}
+
+\section{Avaliação - Prova de Conceito}
+
+Para validar a efetividade da abordagem proposta, foi criado um projeto simples como prova de conceito, utilizando o MTA, o AccTrace e os outros recursos relacionados neste trabalho.
+O objetivo desta prova de conceito foi verificar se o AccTrace:
+
+\begin{itemize}
+  \item Permite a verificação da relação entre os requisitos e casos de uso com a etapa de codificação;
+  \item Permite que seja feito o rastreamento dos requisitos de acessibilidade, desde a sua concepção até as fases de codificação;
+  \item Permite que o desenvolvedor consiga verificar, em nível de código, a associação dos requisitos, modelos UML e técnicas de implementação de acessibilidade.
+\end{itemize}
 
@@
 
-\bibitem{aegis:13} AEGIS Ontology, \url{http://www.aegis-project.eu/index.php?option=com_content&view=article&id=107&Itemid=65}
+\bibitem{aegis:13} AEGIS Ontology, \url{http://www.aegis-project.eu/index.php?option=com_content&view=article&id=107&Itemid=65}. September 2013.
 

### 2014-01-14 paper.tex
.

@@
 
-Para validar a efetividade da abordagem proposta, foi criado um projeto simples como prova de conceito, utilizando o MTA, o AccTrace e os outros recursos relacionados neste trabalho.
+Para validar a efetividade da abordagem proposta, foi criado um projeto simples (software web buscador) como prova de conceito, utilizando o MTA, o AccTrace e os outros recursos relacionados neste trabalho.
 O objetivo desta prova de conceito foi verificar se o AccTrace:
@@
 \end{itemize}
+
+O Desenvolvimento da prova de conceito pode ser encontrada em \cite{branco:13}, e os arquivos e artefatos estão disponíveis para download em \url{http://???}.
+Os objetivos elencados anteriormente foram atingidos: os plugins utilizados são interoperáveis, permitindo a definição e relacionamento dos requisitos e modelos UML, recuperados na etapa de codificação pela ferramenta; o rastreamento pôde ser feito no plugin AccTrace, bem como utilizando a matriz de rastreabilidade gerada pela ferramenta; o desenvolvedor consegue, a partir de um comentário AccTrace personalizado, recuperar informações sobre o requisito e modelo UML associado.
+
+Portanto, foi possível rastrear os requisitos, desde a sua concepção e definição, passando
+pelos modelos e artefatos UML e finalmente, sendo recuperados já na fase de
+codificação. A entrega de informações que possam ser úteis ao desenvolvedor é efetuada.
+O especialista em acessibilidade informa qual técnica de acessibilidade determinado
+modelo/requisito deve usar como guia (seja abordagens, diretrizes, critérios de sucesso,
+técnicas, etc), estando estas informações disponíveis para consulta pelo desenvolvedor. O
+desenvolvedor terá, então, um ponto de partida para codificar o produto da forma correta.
+
+\section{Limitações}
+
+A prova de conceito aqui apresentada visa demostrar como seria a utilização das
+técnicas e ferramentas em um ambiente real. Contudo, não foi possível verificar
+se os passos aqui apresentados seriam realmente utilizados em projeto real. Apesar de estender a
+ISO/IEC 12207, a utilização efetiva do MTA não foi encontrada na
+literatura, para servir como base para este trabalho.
+
+A operação dos plugins construídos/alterados ao longo deste trabalho se
+mostrou difícil em casos pontuais. Algumas dificuldades só foram
+encontradas na execução da prova de conceito, sugerindo que algumas partes
+específicas da ferramenta devem ser adequadas, para que a usabilidade das
+ferramentas seja melhorada.
+
+O volume de dados dos artefatos gerados foi relativamente pequeno. A falta de
+profissionais que efetivamente trabalhem com Engenharia de Requisitos, Modelagem
+e principalmente, a falta de um Especialista em Acessibilidade, profissionais
+estes que poderiam gerar dados mais próximos de um projeto real, impacta
+significativamente na análise do volume de dados \textit{versus} facilidade de
+utilização. As ferramentas geralmente se tornam mais complexas e precisam de
+filtros complexos para entregar os dados corretos no momento certo, escondendo
+os demais, quando o volume de dados é muito grande.
+
+A falta de um Especialista em Acessibilidade impactou negativamente em perceber
+se as informações apresentadas pela ferramenta foram suficientes ou
+desnecessárias. Não foi possível dizer se os mapeamentos encontrados na
+ontologia utilizada foram suficientes para fornecer informações úteis aos
+projetistas/desenvolvedores. Este não é um problema tão sério, já que a
+ontologia pode ser modificada/aprimorada utilizando a ferramenta
+Protegé, devendo o plugin AccTrace ser alterado para que
+comporte a nova realidade do domínio.
+
+Devido a complexidade dos relacionamentos da ontologia utilizada, apenas as
+diretrizes do modelo de referência WCAG 2.0 foi utilizada, por já estarem
+previamente mapeadas. O plugin AccTrace foi inteiramente construído
+utilizando estes relacionamentos como orientação. Por este motivo, a inclusão de
+outros documentos de referência não é uma tarefa simples. Primeiro,
+a ontologia deve ser construída e depois o plugin AccTrace deve ser
+adaptado para refletir a nova realidade do modelo de referência. A consequência
+deste fato é que a atualização dos modelos de referência torna-se uma tarefa
+complexa, inclusive para a atualização do modelo WCAG 2.0, se houver diferenças na estrutura já utilizada na ontologia usada neste trabalho.
+Caso a estrutura entre os elementos da ontologia não seja alterada, não é necessário fazer alterações no plugin AccTrace.
+
+Uma limitação, que deve ser tratada em trabalhos futuros, é que o sistema
+atualmente não trata o caso de remoção de requisitos/modelos UML dos
+artefatos iniciais. Mudanças nos requisitos e modelos UML são automaticamente atualizados (visto que apenas as referências a esses objetos são armazenadas).
+Contudo, as remoções dos elementos nos artefatos causam um comportamento
+inesperado no sistema, obrigando o modelo AccTrace a ser atualizado
+manualmente (com comandos explícitos de remoção das associações no
+plugin).
+
+\section{Conclusões e Trabalhos Futuros}
 
@@
 
+\bibitem{branco:13} Branco, R. G.: Um processo para o desenvolvimento de aplicações Web Acessíveis.
+Master Thesis, UFMS (2013)
+
 \end{thebibliography}

### 2014-01-15 paper.tex
.

@@
 
-O Desenvolvimento da prova de conceito pode ser encontrada em \cite{branco:13}, e os arquivos e artefatos estão disponíveis para download em \url{http://???}.
-Os objetivos elencados anteriormente foram atingidos: os plugins utilizados são interoperáveis, permitindo a definição e relacionamento dos requisitos e modelos UML, recuperados na etapa de codificação pela ferramenta; o rastreamento pôde ser feito no plugin AccTrace, bem como utilizando a matriz de rastreabilidade gerada pela ferramenta; o desenvolvedor consegue, a partir de um comentário AccTrace personalizado, recuperar informações sobre o requisito e modelo UML associado.
+O Desenvolvimento da prova de conceito pode ser encontrada em \cite{branco:13},
+e os arquivos e artefatos estão disponíveis para download em
+\url{https://drive.google.com/file/d/0B9tWGo6C4CtjLWgyQmFkaXVncnM/edit?usp=sharing},
+assim como o código fonte está disponível em
+\url{https://github.com/rodrigogbranco/acctrace}. Os objetivos elencados anteriormente foram atingidos:
+os plugins utilizados são interoperáveis, permitindo a definição e relacionamento dos requisitos e modelos UML, recuperados na etapa de codificação pela ferramenta; o rastreamento pôde ser feito no plugin AccTrace, bem como utilizando a matriz de rastreabilidade gerada pela ferramenta; o desenvolvedor consegue, a partir de um comentário AccTrace personalizado, recuperar informações sobre o requisito e modelo UML associado.
 

### 2014-01-15 paper.tex
.

@@
 \begin{abstract}
-The abstract should summarize the contents of the paper and should
-contain at least 70 and at most 150 words. It should be written using the
-\emph{abstract} environment.
+
+Teste
+
 \keywords{Accessibility, Requirement Traceability, Software Development
 Process, CASE Tool}
-%\keywords{We would like to encourage you to list your keywords within
-%the abstract section}
 \end{abstract}
@@
 
-Os trabalhos anteriores relacionados a este podem ser divididos em três grupos:
-Acessibilidade no Processo de Desenvolvimento de Software, Rastreabilidade de
-Requisitos e Ontologias para o Mapeamento do Domínio.
-
-\subsection{Acessibilidade no Processo de Desenvolvimento de Software}
-
-O tema Acessibilidade no Processo de desenvolvimento de software ainda
-é pouco explorado. Maia \cite{maia:10} propôs  integrar usabilidade e
-acessibilidade aos processos de Engenharia de Software criando o MTA (Modelo de
-Tarefas de Acessibilidade), um modelo de desenvolvimento, baseado na norma ISO/IEC 12207, aplicável
-ao WCAG, que permite o apoio ferramental ao modelo.
-
-\subsection{Rastreabilidade de Requisitos}
-
-Diversos estudos que dizem respeito ao rastreamento de requisitos podem ser
-encontrados na literatura.
-
- Em contrapartida, podem ser encontrados diversos
+Conforme apontado por Dias \cite{analuizadias:2010}, a grande concentração de esforços quando o tema é acessibilidade é em relação a avaliação e validação de produtos, bem como a realização de testes com usuários.
+O tema \textit{Acessibilidade no Processo de desenvolvimento de software} ainda
+é pouco explorado \cite{maia:10}. Em contrapartida, podem ser encontrados diversos
 estudos que dizem respeito ao rastreamento de requisitos
-\cite{5970169,292398,5485417,6405269},
-
-\subsection{Ontologias para o Mapeamento do Domínio}
-
-bem como encontrar estudos que utilizam
+\cite{5970169,292398,5485417,6405269}, bem como encontrar estudos que utilizam
 ontologias para o mapeamento do domínio no processo de desenvolvimento ou na
@@
 não foi encontrado na literatura estudos específicos sobre a rastreabilidade de
-requisitos de acessibilidade durante o processo de desenvolvimento de software.
+requisitos de acessibilidade durante o processo de desenvolvimento de \textit{software}. Também não foi encontrado trabalhos que conseguissem,
+em nível de código, resgatar as informações sobre a rastreabilidade dos requisitos e modelos UML.
 
@@
 
+Este estudo mostrou ser possível especificar, antes das fases de codificação e vinculadas aos modelos e
+requisitos de acessibilidade, as técnicas de implementação que deverão
+ser visualizadas pelos programadores, promovendo a rastreabilidade dos requisitos de acessibilidade durante o processo.
+A utilização de uma ontologia pré-definida do projeto Aegis \cite{aegis:13} ajudou a alcançar tal objetivo, estendendo as
+técnicas de implementação anteriormente ditas para abordagens, diretrizes,
+critérios de sucesso, entre outros.
+
+Assim como no trabalho de Maia \cite{maia:10}, este trabalho pressupõe que um
+especialista em acessibilidade participe do processo de desenvolvimento, principalmente nas
+Subtarefas de acessibilidade propostas pelo MTA. A proposta deste trabalho
+pode se estender para outras áreas (por exemplo, usabilidade, segurança, etc),
+desde que haja um especialista e que o domínio esteja mapeado na forma de uma
+ontologia, ou outro formato que possa ser interpretado por uma ferramenta
+CASE.
+
+As alternativas disponíveis promovem a rastreabilidade entre requisitos \textit{versus} modelos UML.
+Contudo, não foi encontrado como é possível utilizar as alternativas para recuperar informações de rastreabilidade dos requisitos partindo do código fonte do programa.
+Também não foi possível identificar se essas alternativas permitem a associação de técnicas de implementação aos requisitos e modelos UML.
+
+É possível apontar, portanto, cinco grandes contribuições fornecidas por este
+trabalho:
+
+\begin{enumerate}
+  \item estender e dar suporte ao MTA, proposto por \cite{maia:10};
+  \item abordar o rastreamento dos requisitos de acessibilidade em um processo
+  de desenvolvimento que inclui tarefas de acessibilidade (MTA);
+  \item permitir a especificação de técnicas de implementação de acessibilidade, 
+  mapeadas em uma ontologia pré-definida;
+  \item recuperar as informações de rastreabilidade por meio comentários
+  personalizado no código fonte;
+  \item fornecer uma ferramenta como prova de conceito.
+\end{enumerate}
+
+\subsection{Trabalhos Futuros}
+
+É possível identificar várias atividades para trabalhos futuros:
+
+\begin{itemize}
+  \item efetuar um estudo de caso com um projeto real, que utilize o MTA e
+  use os plugins aqui elencados, inclusive os plugins
+  construídos e customizados para promover a rastreabilidade e geração de
+  código;
+  \item efetuar um estudo de caso com usuários reais, utilizando o
+  software construído no item anterior e avaliar a acessibilidade do
+  mesmo;
+  \item estudar a utilização de métodos dinâmicos (a serem definidos) de
+  rastreabilidades dos requisitos;
+  \item efetuar o mapeamento da ontologia para outros documentos de referência
+  em acessibilidade;
+  \item aumentar a usabilidade dos plugins construídos/modificados,
+  melhorando as mensagens apresentadas, aproveitando o relacionamento da
+  ontologia do Projeto Aegis;
+  \item estender o escopo deste trabalho, incluíndo tarefas de testes e
+  integração do software e do sistema (Subtarefas 7, 8, 9 e 10 do
+  MTA);
+  \item tratar o caso de remoção de requisitos/modelos UML já relacionados
+  no modelo AccTrace;
+  \item estender a matriz de rastreabilidade dos requisitos aqui construída,
+  para incluir os casos de testes descritos no item anterior;
+  \item utilizar outro domínio de interesse como base para os estudos futuros,
+  como por exemplo usabilidade de software.
+\end{itemize}
+
 \begin{thebibliography}{4}

### 2014-01-15 paper.tex
.

@@
 
-Teste
+Fornecer produtos web acessíveis é um grande desafio.
+Mecanismos de rastreabilidade de requisitos já existem satisfatoriamente a algum tempo, mas codificar requisitos de acessibilidade ainda é um desafio para muitos desenvolvedores, que podem não possuir as habilidades necessárias para realizar tal tarefa, tendo dificuldade inclusive em interpretar os documentos de referência.
+Neste trabalho é apresentado o AccTrace, uma ferramenta CASE (plugin do Eclipse) que, utilizando uma ontologia de mapeamento do domínio do Projeto Aegis para especificar as técnicas de implementação de acessibilidade, promove a rastreabiliade dos requisitos de acessibilidade desde sua concepção até a fase de codificação, entregando ao desenvolvedor informações úteis para a construção do produto acessível.
 
